---
title: "Case Study: Ankylosing Spondylities"
author: 'Leah Hunt, Ziling Zhang, Jiahui Cheng'
date: 'October 8, 2020'
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
---

```{r Front Matter, include=FALSE}
# clean up & set default chunk options
rm(list = ls())
knitr::opts_chunk$set(echo = FALSE,message=FALSE,warning = FALSE)
# packages
library(tidyverse) 
library(mosaic)
#library(summarytools)
library(kableExtra)
library(magrittr)
library(lme4)
library(lmerTest)
# Data
data <- read.csv("Ankylosing_Spondilitus_data_clean.csv")
colnames(data)[1] <- "patientNo"
```


\newpage


# Project Description

Ankylosing spondylitis (AS) is an inflammatory disease which limits the motion of the spine and muscles. The disease affects the joints of the spinal and peripheral joints such as the shoulder, hip, knee, and ankle. If ribs are affected, it can be difficult to breathe deeply. As a result, people with AS demonstrate inspiratory muscle fatigue during exercise and limited capacity of maximal oxygen. These restrictions lead to decreased daily activity and to decreased quality of life in people with AS.

Some research reveals that exercise is as crucial as drug treatment in the management of AS. To further confirm this hypothesis, a study was carried out at the Royal National Hospital to examine the effects of daily stretching of tissues around the hip on the joints flexibility of AS patients.

In this study, out of 39 patients with AS, 12 were randomly chosen to a control group receving the standard treatment. The other 27 patients received additional stretching excersices. Several measurements were made on each hip when the patients were admitted and three weeks after the treatment. This report will focus on two important measure, flexion and lateral rotation.


## Research Questions


**Question:**   Will treatment be effective in improving flexion and rotation abilities?


## Variables

We analyzed 7 variables collected in the study. Table 1 provides the name and a brief description of each variable. 

```{r fig.cap= "Variables description and measurements"}
# Variable table
variable<-data.frame(
  No = c(1,2,3,4,5,6,7),
  Variable=c("flexionBefore","flexionAfter","rotationBefore","rotationAfter","hip","treatment","patient"),
  Description = 
    c("hip joints flexion before treatment",
      "hip joints flexion after treatment",
      "hip joints rotation before treatment","hip joints rotation after treatment",
      "Right or Left",
      "Control or Stretch",
      "unique patient identification number"), 
  Units = c('degrees','degrees','degrees','degrees','','','1-39'))
knitr::kable(variable,
             caption = "Variable Attributes",
             align = c('c', rep('l', 7)),format="latex") %>%
kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
                          font_size = 7, latex_options = "HOLD_position")
```


# Exploratory Data Analysis (EDA)

There were no clear errors or missing values in the dataset as presented. 

The descriptive statistics will show the summary statistics including min, max, median, mean, and variance for both control and treatment groups including both right and left hips because from the visulization, it seems that there is no huge difference between right and left hips. The complete descriptive statistics of all control variables is in the appendix. 

*Table 2* is the descriptive statistics for degrees of motions of flexion and rotation if no streching exercises are applied. For the *SAM*, both rotation and flexion degrees have slight improvement, while flexion seems to more improvement. For the *SAV*, there is no obvious fluctuation in either variable. 

```{r fig.cap="The table displays descriptive statistics for flexion and rotation in control group"}
data$flexiondiff=data$flexionAfter-data$flexionBefore
data$rotationdiff=data$rotationAfter-data$rotationBefore
dt<-tidyr::pivot_longer(data[1:24,], 5:10, names_to="Stat")
stats <- psych::describeBy(
  x = as.numeric(dt$value), group= dt$Stat,
  na.rm = TRUE, interp = TRUE, 
  quant= c(0.25,0.75), skew = TRUE, 
  digits = 2, mat = TRUE)
df <- data.frame(
  Var = stats$group1, 
  `Sample Min` = stats$min, 
  `Sample Median` = stats$median, 
  `Sample Max` = stats$max, 
  SAM = stats$mean, 
  SAV = round((stats$sd)^2,3))
knitr::kable(df, escape=F,
             caption = "Statistics in Control Group",
             align = c('c', rep('r', 9))) %>%
kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
                          font_size = 9, latex_options = "HOLD_position") 
```

*Table 3* shows the summary statistics for treatment group, after streching exercises are applied. For *SAM*, both abilities have degrees improvements of around 7 degrees. Thus, it seems like exercises are effective. 

One thing notice from both control and treatment group is the *SAV* for flexion and rotation. Both behind and after flexion variables have larger *SAV* than rotation. Thus, flexion abilities seem to have more fluctuation than rotation. 
```{r fig.cap="The table displays descriptive statistics for flexion and rotation in treatment group"}
data$flexiondiff=data$flexionAfter-data$flexionBefore
data$rotationdiff=data$rotationAfter-data$rotationBefore
dt<-tidyr::pivot_longer(data[25:78,], 5:10, names_to="Stat")
stats <- psych::describeBy(
  x = as.numeric(dt$value), group= dt$Stat,
  na.rm = TRUE, interp = TRUE, 
  quant= c(0.25,0.75), skew = TRUE, 
  digits = 2, mat = TRUE)
df <- data.frame(
  Var = stats$group1, 
  `Sample Min` = stats$min, 
  `Sample Median` = stats$median, 
  `Sample Max` = stats$max, 
  SAM = stats$mean, 
  SAV = round((stats$sd)^2,3))
knitr::kable(df, escape=F,
             caption = "Statistics in Treatment Group",
             align = c('c', rep('r', 9))) %>%
kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
                          font_size = 9, latex_options = "HOLD_position") 
```
 

Histogram of the density plot can better shows the distribution of both before and after performance for both abilities. *Figure 1* is the histogram of flexion in control and treatment group. The x axis is the performance in degrees. The y axis is the density. For control group, corresponding to summary statistics above, the means of flexion before and after are around 110 degrees. For treatment group, the mean of flexion after seems to be slightly larger than mean of flexion before. 

```{r fig.height=3,fig.width=10,fig.cap="Histogram of flexion"}
data %>%
  gather("key","value",5:6) %>%
  filter(key %in% c("flexionAfter", "flexionBefore")) %>%
  ggplot(aes(x=value,fill=key)) + 
  geom_histogram(aes(y = ..density..),alpha=0.3,colour="grey40",bins=20) +
  geom_density(colour="grey40",alpha=0.1) +
  facet_grid(.~treatment) + 
  xlab("flexion")
```


\newpage




*Figure 2* shows the similar histogram for rotation. In general, rotation has less degrees in motions than flexion with range 0 to 50. For both control and treatment groups, there is no obvious difference of rotation after and rotation before. 

```{r fig.height=3,fig.width=10,fig.cap="Histogram of rotation"}
# Figure 2
data %>%
  gather("key","value",7:8) %>%
  filter(key %in% c("rotationAfter", "rotationBefore")) %>%
  ggplot(aes(x=value,fill=key)) + 
  geom_histogram(aes(y = ..density..),alpha=0.3,colour="grey40",bins=20) +
  geom_density(colour="grey40",alpha=0.1) +
  facet_grid(.~treatment) + 
  xlab("rotation")
```




After examing the distribution of measurement, it seems appropriate to use the difference of measurement after and before the treatment as an indicator of the improvement of the patients. We will use two variables *flexionIncrease* and *rotationIncrease* to denote the increase for the two measurements.
Figure 3 and 4 shows the comparasion of the increase on left and right hips and on control and stretch groups.

*Figure 3* is the boxplot for flexionIncrease. It seems that there are some potantial outliers that falls out of range. However, considering flexionIncrease is a calculated data and both flexionBefore and flexionAfter don't show typical outliers, we will still count these points into considerarion. The boxplot shows that there is no obvious difference between left and right hip, while left hip control group might have larger fluctuation. 

```{r fig.height=3,fig.weight=10,fig.cap="boxplot of flextion increase"}
# Figure 3
data %>%
  mutate(flexionIncrease = flexionAfter-flexionBefore, 
         rotationIncrease = rotationAfter-rotationBefore) %>%
  ggplot(aes(y=flexionIncrease, x=hipRL, fill=treatment)) +
  geom_boxplot()
```


*Figure 4* shows similar boxplot for rotation. Here, it seems that treatment group might have better performance in motions degrees. Similar to *Figure 3*, there is no obvious difference between both hips.

```{r fig.height=3,fig.weight=10,fig.cap="boxplot of rotation increase"}
# Figure 4
data %>%
  mutate(flexionIncrease = flexionAfter-flexionBefore, 
         rotationIncrease = rotationAfter-rotationBefore) %>%
  ggplot(aes(y=rotationIncrease, x=hipRL, fill=treatment)) +
  geom_boxplot()
```







# Statistical Analysis 

As described in previous sections, this dataset actually contains three different levels of observations. Level 1 is the record of each hip. Level 2 consists of different patients, and level 3 contains our two treatment groups. In order to take into consideration the variation among patients and treatment groups, we will use a three-level model to represent our problem.

Then we also need to choose which effects we should include in our model for different sources of variability. Since the treatment group is our research goal, we will associate fixed effects to the treatment. And because our patients are randomly chosen, we assume the patients are from certain probability distribution and associate random effects with them.


## Model Structure

We will build a multi mixed-effect linear model for this project. The response variables will be *flexionAfter* and *rotationAfter* for both abilities. The independent variables include *flexionBefore* adn *rotationBefore* as random effects variables, *hip* (right or left) and *treatment* (control or stretch) are fixed effects variables. *patientNo* will also be random effects variable since for each patient there is right hip data and left hip data. The order of observation is also not independent. 

The two level model can be expressed as

-Level One (Composite model)

$$flexionAfter_{ij} = \beta_1*flexionBefore_{ij} + \beta_2*hip + \beta_3*treatment + \beta_{4_i} + \epsilon_{ij} \space \text{where}\space \epsilon_{ij}\sim N(0,\sigma_{\epsilon}) $$
-Level Two (Effects on PatientNo)

$$\beta_{4_i} = \alpha_i +  \mu_i\space\text{where}\space \epsilon_{i}\sim N(0,\sigma_{\mu})$$
$$\alpha_i\space\text{is the distinct observation we have}$$
Notes: hip = 0 if it is a left hip and 1 if it is a right hip. Treatment = 0 if it is the control and 1 if it is the treatment group, i represents the patient number and j represents which hip is being considered



## Model results

*Table 4 and 5* shows the flexion model. The coefficient of treatment is 7.8. It means if all other variables including hip, patient are constant, the stretching exercises can improve flexion abilities by around 7.8 degrees. The p-value for treatment is close to 0. Thus, the treatment is effective in improving flexion abilities. 

Moreover, the coefficient of hip is -0.23. Thus, right hip tend to have slightly worse performance than left hip if other variables constant. The high p-value shows that hip might not be an issue in determining improvement. 

The variances of patientsNo and flexionBefore are 25, 63, repesctively. Considering the total variation is around 60, the random effects here are reasonable. 

```{r Model1-flexion} 
data$patientNo <- as.factor(data$patientNo)
model1Flexion <- lmer(flexionAfter ~ (1 | flexionBefore) + treatment + 
                        hipRL + (1 | patientNo), data = data)
mod <- summary(model1Flexion)$coefficients
rownames(mod) <- c("(Intercept)", "Treatment", "Right Hip")
knitr::kable(mod, escape=F,
             align = c('c', rep('r', 9)),
             caption = "Effects on Flexion Improvement") %>%
  kableExtra::kable_styling(latex_options = "HOLD_position") 
m <- summary(model1Flexion)
n <- m$varcor
df <- data.frame(Group = c("Patient", "Flexion Before", "Residual"), 
                 Variance = c(25.999, 63.097, 8.875))
#n$Std.Dev. <- as.numeric(n$Std.Dev.)^2
knitr::kable(df, escape=F,
             align = c('c', rep('r', 9)), caption = "Variance for Random Effects of Flexion") %>%
  kableExtra::kable_styling(latex_options = "HOLD_position") 
```
 

Similarily, *table 6 and 7* show the rotation model. The coefficient of treatment is 2.3. It means if all other variables including hip, patient are constant, the stretching exercises can improve rotation abilities by around 2.3 degrees. The p-value for treatment is close to 0. Thus, the treatment is effective in improving rotation abilities. However, cocmpared to flexion, the improvement in rotation is lighter.

Moreover, the coefficient of hip is 5.8. Thus, if other variables contant, right hip tends to increase 5.8 more degrees than left hip. The close-to-0 p value also shows that both treatment and hip are effective in rotation performance. 

The variances of patientsNo and rotationBefore are 23, 65, repesctively. Considering the total variation is around 60, the random effects here are reasonable. 

 
 
```{r Model1-rotation} 
model1Rotation <- lmer(rotationAfter ~ (1 | rotationBefore) + hipRL + 
                         treatment + (1 | patientNo), data = data)
mod <- summary(model1Rotation)$coefficients
rownames(mod) <- c("(Intercept)", "Treatment", "Right Hip")
# knitr::kable(mod, escape=F,
#              align = c('c', rep('r', 9)),
#              caption = "Effects on Rotation Improvement") %>%
#   kableExtra::kable_styling(position ="float_left", full_width = F) 
m <- summary(model1Rotation)
n <- m$varcor
df <- data.frame(Group = c("Patient", "Rotation Before", "Residual"), Variance = c(23.218, 65.006, 7.807))
#n$Std.Dev. <- as.numeric(n$Std.Dev.)^2
#gridExtra::grid.arrange(
knitr::kable(mod, escape=F,
             align = c('c', rep('r', 9)),
             caption = "Effects on Rotation Improvement") %>%
  kableExtra::kable_styling(latex_options = "HOLD_position")
knitr::kable(df, escape=F,
             align = c('c', rep('r', 9)), , caption = "Variance for Random Effects of Rotation") %>%
  kableExtra::kable_styling(latex_options = "HOLD_position") 
#nrow = 1
#)
```


As a conclusion, the variation of slope and coefficients of randome effects of both models are shown above, adding only fixed effects, we have

$$Predicted\space flexionAfter_{ij} = 114.7273+\beta_1*flexionBefore_{ij} -0.2293*hip + 7.8325*treatment + \beta_{4_i}  $$
 $$Predicted\space rotationAfter_{ij} = 23.064+\beta_1*rotationBefore_{ij} +2.88*hip + 5.834*treatment + \beta_{4_i}  $$




# Recommendations 

**Research Question:**  Will treatment be effective in improving flexion and rotation abilities?

The treatment seems to be effective in both flexion and rotation abilities from the model resutls. Moreover, the hip also matters in rotation abilities. Thus, stretching exercises show effective results in helping patients recover from Ankulosing Sponsilitus. 

For further research, we will recommen the client to look for other effects of stretching exercises such as other motions in addition to flexion and rotation. The client can also investigate the relationship between hips and spine abilities. 


# Additional Considerations and limitations

- Some other factors that may also affect the improvement like patients age, years of getting AS, health condition etc. are not included in the dataset. Those information will be helpful to further distinguish the reason of improvement.
- the observation here is not totally independent. The order of observation might be affected by areas and relationships between patients. Thus, the client might want to perform similar research again with randomly selected patients.

# Resources

http://www.lowtem.hokudai.ac.jp/plantecol/akihiro/obenkyou/GLMMexample_En.pdf

https://bookdown.org/roback/bookdown-BeyondMLR/ch-multilevelintro.html

https://psu.mediaspace.kaltura.com/media/STAT+470W--Ankylosing+spondylitis+interview+with+Dr.+Elizabeth+Nemenz/1_n2v1u1p8


\newpage


# Technical Appendix

## A: Assumption checks

Generally, multilevel models response is assumed to be linear with respect to the predicted value with constant variance. Residual plots are a useful tool to examine these assumptions. Our model also assumes that the residuals follow the Normal distribution. We can use Q-Q plot to test the normality of model residuals. From QQ plots, the residuals of the two models are close to normal distributions.Overall, the results indicate that our model meets the basic assumptions.

```{r fig.height=3,fig.weight=10,fig.cap="Residuals plot for models"}
par(mfrow=c(2,2))
gridExtra::grid.arrange(
ggplot(fortify.merMod(model1Flexion), aes(x = .fitted, y = .resid)) + 
  geom_point() +
  geom_hline(yintercept = 0) +
  ylim(-11,17) +
  ggtitle("residuals of flexion model")
,
ggplot(fortify.merMod(model1Rotation), aes(x = .fitted, y = .resid)) + 
  geom_point() +
  geom_hline(yintercept = 0) +
  ylim(-11,17) +
  ggtitle("residuals of rotation model")
,
nrow = 1)
```




```{r fig.height=3,fig.weight=10,fig.cap="Q-Q plot for models"}
gridExtra::grid.arrange(
ggplot(fortify.merMod(model1Flexion), aes(sample = .resid)) + 
  stat_qq() + 
  stat_qq_line() +
  ylim(-10,10) +
  ggtitle("Q-Q plot of flexion model")
,
ggplot(fortify.merMod(model1Rotation), aes(sample = .resid)) + 
  stat_qq() + 
  stat_qq_line()+
  ylim(-10,10) +
  ggtitle("Q-Q plot of rotation model")
,
nrow = 1)
```

\newpage

## B Correlation plot for flexionIncrease and RotationIncrease on both hips


We will look at the correlations among patients' left and right hips. It is natural to think that the improvement on the same patients should be positively correlated. The following figure further confirms this relationship. 

```{r EDA-corrplot, fig.cap="scatterplot of increase in left and right hips"}
# Figure 8
data <- data %>%
  mutate(flexionIncrease = flexionAfter-flexionBefore, 
         rotationIncrease = rotationAfter-rotationBefore)
gridExtra::grid.arrange(
data %>%
  select(treatment, patientNo, flexionIncrease, hipRL) %>%
  group_by(patientNo) %>%
  spread(hipRL, flexionIncrease) %>%
  ggplot(aes(x=left, y=right)) +
  geom_smooth(method = 'loess', alpha=0.1, span=1, linetype = "dashed")+
  geom_point(aes(color=treatment)) +
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = 0) +
  ggtitle("flexion Increase")
,
data %>%
  select(treatment, patientNo, rotationIncrease, hipRL) %>%
  group_by(patientNo) %>%
  spread(hipRL, rotationIncrease) %>%
  ggplot(aes(x=left, y=right)) +
  geom_smooth(method = 'loess',alpha=0.1, span=1, linetype = "dashed")+
  geom_point(aes(color=treatment)) +
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = 0) +
  ggtitle("ratation Increase")
,nrow = 1)
```
\newpage

## C Scatter plot for flexionIncrease and RotationIncrease 

The vertical distance from each point to the diagonal line `y = x` represents the change of measurements after treatment. Above the line means an inprovement whereas below the line means a worse condition. These two figures also show that the stretch group has better improvement than the control group since the fitted curve for the stretch group is above the curve for the control group. 


```{r fig.height=3,fig.weight=10,fig.cap="Scatter plot of flexion before and after treatment"}

data %>%
  ggplot(aes(x = flexionBefore, y = flexionAfter)) + 
  geom_point(aes(color = treatment)) + 
  geom_smooth(aes(color = treatment),method = "loess", se = 0) + 
  geom_abline(slope = 1)
```

```{r fig.height=3,fig.weight=10,fig.cap="Scatter plot of rotation before and after treatment"}

data %>%
  ggplot(aes(x = rotationBefore, y = rotationAfter)) + 
  geom_point(aes(color = treatment)) + 
  geom_smooth(aes(color = treatment),method = "loess", se = 0) + 
  geom_abline(slope = 1)
```
\newpage

## D Posthoc Analysis on Bonferroni Confidence Interval

```{r}
intervals <-confint(model1Flexion, level = 0.95, oldNames = FALSE)
row.names(intervals)[c(4,1,2)] <- c("Intercept", "Patient Number", "Rotation Before")
knitr::kable(format = "latex",
  intervals[c(4,1,2),],
   digits = 3,
   caption = "Upper and Lower Confidence Bounds-90% SCI, Bonferroni Adj",
 align = c('l',rep('c',2))
) %>%
  kableExtra::kable_styling(
     bootstrap_options = c("striped", "condensed"),
    font_size = 12, latex_options = "HOLD_position")

intervals <-confint(model1Rotation, level = 0.95, oldNames = FALSE)
 row.names(intervals)[c(4,1,2)] <- c("Intercept", "Patient Number", "Rotation Before")
 knitr::kable(format="latex",
  intervals[c(4,1,2),],
  digits = 3,
   caption = "Upper and Lower Confidence Bounds-90% SCI, Bonferroni Adj",
 align = c('l',rep('c',2))
) %>%
  kableExtra::kable_styling(
     bootstrap_options = c("striped", "condensed"),
    font_size = 12, latex_options = "HOLD_position")
```


### R Script
```{r codeAppendix, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}
# Reprinted code chunks used previously for analysis
```
