---
title: "EDA_Case_Study2"
author: "Leah Hunt"
date: "10/8/2020"
output: pdf_document
---

```{r}
library(ggplot2)
```


```{r}
data <- read.table("tidyAS.txt")
head(data)
```

Think through the five steps in designing an experiment and reflect on how they apply to the Ankylosing Spondylitis study:
* What is/are the research hypothesis/es?
null: stretching does nothing
alternative: stretching improves "condition"
~split "condition into flexion and rotation

* What are the key outcome, explanatory, and nuisance variables?  Can you name 2 reasonable outcome variables?
~Outcome: flexion and rotation
~explanatory: treatment
~nuisance: before, left/right hip

*What factors might have led to the particular choices of how many subjects to use?
~Likely chosen by people that had condition and that were available for study 

*What do you know about how subjects were assigned to conditions?
~All we know is that there are twice as many stretching as control; implied random assignment but no description given

*What are the analyses that will address the hypotheses? In particular, what sources of variation will you include in your model and should they be fixed or random?
~ANOVA, mixed effects

*What types of graphical displays might be helpful at both the exploratory data analysis and the final analysis phases?
~ Numerical summaries comparing left and right hip
  ~ General numerical summaries too (maybe something with improvement)
~ points with left and right and before and after for flexion and rotation (put line between control and treatment)

```{r}
colors <- c("Left Hip" = "blue", "Right Hip" = "red")

plot<-ggplot2::ggplot(aes(x = patientNo), data= data) +
  geom_point(aes(x = patientNo + .1, y = flexionBefore), color = "red", data = data[data$hip == "Right",]) + 
        geom_point(aes(x = patientNo + .1, y = flexionAfter), color = "red", data = data[data$hip == "Right",]) +
        geom_point(aes(x = patientNo - .1, y = flexionBefore), color = "blue", data = data[data$hip == "Left",]) + 
        geom_point(aes(x = patientNo - .1, y = flexionAfter), color = "blue", data = data[data$hip == "Left",]) +
  geom_segment(aes(x = patientNo + .1, xend = patientNo + .1, y = flexionBefore, yend = flexionAfter, color = "Right Hip"), data = data[data$hip == "Right",], arrow = arrow(length = unit(0.1, "inches")))+
   geom_segment(aes(x = patientNo - .1, xend = patientNo - .1, y = flexionBefore, yend = flexionAfter, color = "Left Hip"), data = data[data$hip == "Left",], arrow = arrow(length = unit(0.1, "inches")))+
        geom_vline(xintercept = 12.5) +
  labs(x = "Patient",
         y = 'Flexion Score',
         color = "Legend") +
        ggtitle("Flexion Changes for Left and Right Hips")+
  
    scale_color_manual(values = colors) +
        theme(axis.text = element_text(size=10),
              plot.title = element_text(size=10, face="bold",hjust=0.5),
              axis.title = element_text(size=10),
              panel.background = element_rect(fill = "white", color="black"),
              legend.text = element_text(size=14),
              legend.title = element_text(size=16)
        )

plot

```
Yes, I know I said no arrows, but then I realized jitter made it work better and this is actually a much more readable visualization. 


```{r}
colors <- c("Left Hip" = "blue", "Right Hip" = "red")

plot<-ggplot2::ggplot(aes(x = patientNo), data= data) +
  geom_point(aes(x = patientNo + .1, y = rotationBefore), color = "red", data = data[data$hip == "Right",]) + 
        geom_point(aes(x = patientNo + .1, y = rotationAfter), color = "red", data = data[data$hip == "Right",]) +
        geom_point(aes(x = patientNo - .1, y = rotationBefore), color = "blue", data = data[data$hip == "Left",]) + 
        geom_point(aes(x = patientNo - .1, y = rotationAfter), color = "blue", data = data[data$hip == "Left",]) +
  geom_segment(aes(x = patientNo + .1, xend = patientNo + .1, y = rotationBefore, yend = rotationAfter, color = "Right Hip"), data = data[data$hip == "Right",], arrow = arrow(length = unit(0.1, "inches")))+
   geom_segment(aes(x = patientNo - .1, xend = patientNo - .1, y = rotationBefore, yend = rotationAfter, color = "Left Hip"), data = data[data$hip == "Left",], arrow = arrow(length = unit(0.1, "inches")))+
        geom_vline(xintercept = 12.5) +
  labs(x = "Patient",
         y = 'Rotation Score',
         color = "Legend") +
        ggtitle("Rotation Changes for Left and Right Hips")+
  
    scale_color_manual(values = colors) +
        theme(axis.text = element_text(size=10),
              plot.title = element_text(size=10, face="bold",hjust=0.5),
              axis.title = element_text(size=10),
              panel.background = element_rect(fill = "white", color="black"),
              legend.text = element_text(size=14),
              legend.title = element_text(size=16)
        )

plot

```
Notes:
*Looks like rotation has more distinct difference in amount of change between treatment and non-treatment
*Left hips may be slightly more prone to decrease? Especially in control?


Below are some things that aren't necessarily necessary to show/present but just exploring how related flexion and rotation are related
```{r}
cor(data$flexionAfter, data$rotationAfter)
cor(data$flexionBefore, data$rotationBefore)
cor(data$flexionAfter-data$flexionBefore, data$rotationAfter-data$rotationBefore)
```
```{r}
plot(data$flexionAfter, data$rotationAfter)
plot(data$flexionBefore, data$rotationBefore)
plot(data$flexionAfter-data$flexionBefore, data$rotationAfter-data$rotationBefore)
```


