---
title: "EDA_Case_Study2"
author: "Leah Hunt"
date: "10/8/2020"
output: pdf_document
---

```{r}
library(ggplot2)
library(tidyr)
```


```{r}
data <- read.csv("Ankylosing_Spondilitus_data_clean.csv")
colnames(data)[1] <- "patientNo"
head(data)
```

Think through the five steps in designing an experiment and reflect on how they apply to the Ankylosing Spondylitis study:
* What is/are the research hypothesis/es?
null: stretching does nothing
alternative: stretching improves "condition"
~split "condition into flexion and rotation

* What are the key outcome, explanatory, and nuisance variables?  Can you name 2 reasonable outcome variables?
~Outcome: flexion and rotation
~explanatory: treatment
~nuisance: before, left/right hip

*What factors might have led to the particular choices of how many subjects to use?
~Likely chosen by people that had condition and that were available for study 

*What do you know about how subjects were assigned to conditions?
~All we know is that there are twice as many stretching as control; implied random assignment but no description given

*What are the analyses that will address the hypotheses? In particular, what sources of variation will you include in your model and should they be fixed or random?
~ANOVA, mixed effects

*What types of graphical displays might be helpful at both the exploratory data analysis and the final analysis phases?
~ Numerical summaries comparing left and right hip
  ~ General numerical summaries too (maybe something with improvement)
~ points with left and right and before and after for flexion and rotation (put line between control and treatment)

```{r}
colors <- c("Left Hip" = "blue", "Right Hip" = "red")

plot<-ggplot2::ggplot(aes(x = patientNo), data= data) +
  geom_point(aes(x = patientNo + .1, y = flexionBefore), color = "red", data = data[data$hipRL == "right",]) + 
        geom_point(aes(x = patientNo + .1, y = flexionAfter), color = "red", data = data[data$hipRL == "right",]) +
        geom_point(aes(x = patientNo - .1, y = flexionBefore), color = "blue", data = data[data$hipRL == "left",]) + 
        geom_point(aes(x = patientNo - .1, y = flexionAfter), color = "blue", data = data[data$hipRL == "left",]) +
  geom_segment(aes(x = patientNo + .1, xend = patientNo + .1, y = flexionBefore, yend = flexionAfter, color = "Right Hip"), data = data[data$hipRL == "right",], arrow = arrow(length = unit(0.1, "inches")))+
   geom_segment(aes(x = patientNo - .1, xend = patientNo - .1, y = flexionBefore, yend = flexionAfter, color = "Left Hip"), data = data[data$hipRL == "left",], arrow = arrow(length = unit(0.1, "inches")))+
        geom_vline(xintercept = 12.5) +
  labs(x = "Patient",
         y = 'Flexion Score',
         color = "Legend") +
        ggtitle("Flexion Changes for Left and Right Hips")+
  
    scale_color_manual(values = colors) +
        theme(axis.text = element_text(size=10),
              plot.title = element_text(size=10, face="bold",hjust=0.5),
              axis.title = element_text(size=10),
              panel.background = element_rect(fill = "white", color="black"),
              legend.text = element_text(size=10),
              legend.title = element_text(size=10)
        )

plot

```
Yes, I know I said no arrows, but then I realized jitter made it work better and this is actually a much more readable visualization. 


```{r}
colors <- c("Left Hip" = "blue", "Right Hip" = "red")

plot<-ggplot2::ggplot(aes(x = patientNo), data= data) +
  geom_point(aes(x = patientNo + .1, y = rotationBefore), color = "red", data = data[data$hipRL == "right",]) + 
        geom_point(aes(x = patientNo + .1, y = rotationAfter), color = "red", data = data[data$hipRL == "right",]) +
        geom_point(aes(x = patientNo - .1, y = rotationBefore), color = "blue", data = data[data$hipRL == "left",]) + 
        geom_point(aes(x = patientNo - .1, y = rotationAfter), color = "blue", data = data[data$hipRL == "left",]) +
  geom_segment(aes(x = patientNo + .1, xend = patientNo + .1, y = rotationBefore, yend = rotationAfter, color = "Right Hip"), data = data[data$hipRL == "right",], arrow = arrow(length = unit(0.1, "inches")))+
   geom_segment(aes(x = patientNo - .1, xend = patientNo - .1, y = rotationBefore, yend = rotationAfter, color = "Left Hip"), data = data[data$hipRL == "left",], arrow = arrow(length = unit(0.1, "inches")))+
        geom_vline(xintercept = 12.5) +
  labs(x = "Patient",
         y = 'Rotation Score',
         color = "Legend") +
        ggtitle("Rotation Changes for Left and Right Hips")+
  
    scale_color_manual(values = colors) +
        theme(axis.text = element_text(size=10),
              plot.title = element_text(size=10, face="bold",hjust=0.5),
              axis.title = element_text(size=10),
              panel.background = element_rect(fill = "white", color="black"),
              legend.text = element_text(size=10),
              legend.title = element_text(size=10)
        )

plot

```
Notes:
*Looks like rotation has more distinct difference in amount of change between treatment and non-treatment
*Left hips may be slightly more prone to decrease? Especially in control?


```{r}
data2 <- data
data2$deltaRotation <- data2$rotationAfter - data2$rotationBefore
data2$deltaFlexion <- data2$flexionAfter - data2$flexionBefore
dt<-tidyr::pivot_longer(data2, 9:10, names_to="Stat")
stats <- psych::describeBy(
  x = as.numeric(dt$value), group= dt$Stat,
  na.rm = TRUE, interp = TRUE, 
  quant= c(0.25,0.75), skew = TRUE, 
  digits = 2, mat = TRUE)
df1 <- data.frame(
  Treatment = c("Combined", "Combined"),
  Hip = c("Combined", "Combined"),
  Var = stats$group1, 
  `Sample Min` = stats$min, 
  `Sample Median` = stats$median, 
  `Sample Max` = stats$max, 
  SAM = stats$mean, 
  SAV = round((stats$sd)^2,3), 
  `Sample Skewness` = stats$skew)
#df <- df[c(1,5:17, 20),]
#df$SAV=kableExtra::cell_spec(df$SAV, background  = ifelse(df$SAV > 70, "yellow", "white"))

knitr::kable(df, escape=F,
             caption = "Statistics by Variable",
             align = c('c', rep('r', 9))) %>%
kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
                          font_size = 9, latex_options = "HOLD_position") 
```

```{r}
dt<-tidyr::pivot_longer(data2[data2$hipRL == "left",], c(9:10), names_to="Stat")
stats <- psych::describeBy(
  x = as.numeric(dt$value), group= dt$Stat,
  na.rm = TRUE, interp = TRUE, 
  quant= c(0.25,0.75), skew = TRUE, 
  digits = 2, mat = TRUE)
df2 <- data.frame(
  Treatment = c("Combined", "Combined"),
  Hip = c("Left", "Left"),
  Var = stats$group1, 
  `Sample Min` = stats$min, 
  `Sample Median` = stats$median, 
  `Sample Max` = stats$max, 
  SAM = stats$mean, 
  SAV = round((stats$sd)^2,3), 
  `Sample Skewness` = stats$skew)
#df <- df[c(1,5:17, 20),]
#df$SAV=kableExtra::cell_spec(df$SAV, background  = ifelse(df$SAV > 70, "yellow", "white"))

knitr::kable(df, escape=F,
             caption = "Statistics by Variable for left hip",
             align = c('c', rep('r', 9))) %>%
kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
                          font_size = 9, latex_options = "HOLD_position") 
```


```{r}
dt<-tidyr::pivot_longer(data2[data2$hipRL == "right",], c(9:10), names_to="Stat")
stats <- psych::describeBy(
  x = as.numeric(dt$value), group= dt$Stat,
  na.rm = TRUE, interp = TRUE, 
  quant= c(0.25,0.75), skew = TRUE, 
  digits = 2, mat = TRUE)
df3 <- data.frame(
  Treatment = c("Combined", "Combined"),
  Hip = c("Right", "Right"),
  Var = stats$group1, 
  `Sample Min` = stats$min, 
  `Sample Median` = stats$median, 
  `Sample Max` = stats$max, 
  SAM = stats$mean, 
  SAV = round((stats$sd)^2,3), 
  `Sample Skewness` = stats$skew)


# knitr::kable(df, escape=F,
#              caption = "Statistics by Variable for right hip",
#              align = c('c', rep('r', 9))) %>%
# kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
#                           font_size = 9, latex_options = "HOLD_position") 
```

```{r}
data3 <- data2[data2$hipRL == "right",]
data3 <- data3[data2$treatment == "control",]
dt<-tidyr::pivot_longer(data3, c(9:10), names_to="Stat")
stats <- psych::describeBy(
  x = as.numeric(dt$value), group= dt$Stat,
  na.rm = TRUE, interp = TRUE, 
  quant= c(0.25,0.75), skew = TRUE, 
  digits = 2, mat = TRUE)
df4 <- data.frame(
  Treatment = c("Control", "Control"),
  Hip = c("Right", "Right"),
  Var = stats$group1, 
  `Sample Min` = stats$min, 
  `Sample Median` = stats$median, 
  `Sample Max` = stats$max, 
  SAM = stats$mean, 
  SAV = round((stats$sd)^2,3), 
  `Sample Skewness` = stats$skew)


knitr::kable(df, escape=F,
             caption = "Statistics by Variable for right hip",
             align = c('c', rep('r', 9))) %>%
kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
                          font_size = 9, latex_options = "HOLD_position") 
```

```{r}
data3 <- data2[data2$hipRL == "right",]
data3 <- data3[data2$treatment == "treatment",]

dt<-tidyr::pivot_longer(data3, c(9:10), names_to="Stat")

stats <- psych::describeBy(
  x = as.numeric(dt$value), group= dt$Stat,
  na.rm = TRUE, interp = TRUE, 
  quant= c(0.25,0.75), skew = TRUE, 
  digits = 2, mat = TRUE)
df5 <- data.frame(
  Treatment = c("Treatment", "Treatment"),
  Hip = c("Right", "Right"),
  Var = stats$group1, 
  `Sample Min` = stats$min, 
  `Sample Median` = stats$median, 
  `Sample Max` = stats$max, 
  SAM = stats$mean, 
  SAV = round((stats$sd)^2,3), 
  `Sample Skewness` = stats$skew)


knitr::kable(df, escape=F,
             caption = "Statistics by Variable for right hip",
             align = c('c', rep('r', 9))) %>%
kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
                          font_size = 9, latex_options = "HOLD_position") 
```

```{r}
data3 <- data2[data2$hipRL == "left",]
data3 <- data3[data2$treatment == "control",]
dt<-tidyr::pivot_longer(data3, c(9:10), names_to="Stat")
stats <- psych::describeBy(
  x = as.numeric(dt$value), group= dt$Stat,
  na.rm = TRUE, interp = TRUE, 
  quant= c(0.25,0.75), skew = TRUE, 
  digits = 2, mat = TRUE)
df6 <- data.frame(
  Treatment = c("Control", "Control"),
  Hip = c("Left", "Left"),
  Var = stats$group1, 
  `Sample Min` = stats$min, 
  `Sample Median` = stats$median, 
  `Sample Max` = stats$max, 
  SAM = stats$mean, 
  SAV = round((stats$sd)^2,3), 
  `Sample Skewness` = stats$skew)


knitr::kable(df, escape=F,
             caption = "Statistics by Variable for right hip",
             align = c('c', rep('r', 9))) %>%
kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
                          font_size = 9, latex_options = "HOLD_position") 
```

```{r}
data3 <- data2[data2$hipRL == "left",]
data3 <- data3[data2$treatment == "treatment",]

dt<-tidyr::pivot_longer(data3, c(9:10), names_to="Stat")

stats <- psych::describeBy(
  x = as.numeric(dt$value), group= dt$Stat,
  na.rm = TRUE, interp = TRUE, 
  quant= c(0.25,0.75), skew = TRUE, 
  digits = 2, mat = TRUE)
df7 <- data.frame(
  Treatment = c("Treatment", "Treatment"),
  Hip = c("Left", "Left"),
  Var = stats$group1, 
  `Sample Min` = stats$min, 
  `Sample Median` = stats$median, 
  `Sample Max` = stats$max, 
  SAM = stats$mean, 
  SAV = round((stats$sd)^2,3), 
  `Sample Skewness` = stats$skew)


knitr::kable(df, escape=F,
             caption = "Statistics by Variable for right hip",
             align = c('c', rep('r', 9))) %>%
kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
                          font_size = 9, latex_options = "HOLD_position") 
```

```{r}
```{r}
dt<-tidyr::pivot_longer(data2[data2$hipRL == "right",], c(9:10), names_to="Stat")
stats <- psych::describeBy(
  x = as.numeric(dt$value), group= dt$Stat,
  na.rm = TRUE, interp = TRUE, 
  quant= c(0.25,0.75), skew = TRUE, 
  digits = 2, mat = TRUE)
df3 <- data.frame(
  Treatment = c("Combined", "Combined"),
  Hip = c("Right", "Right"),
  Var = stats$group1, 
  `Sample Min` = stats$min, 
  `Sample Median` = stats$median, 
  `Sample Max` = stats$max, 
  SAM = stats$mean, 
  SAV = round((stats$sd)^2,3), 
  `Sample Skewness` = stats$skew)


# knitr::kable(df, escape=F,
#              caption = "Statistics by Variable for right hip",
#              align = c('c', rep('r', 9))) %>%
# kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
#                           font_size = 9, latex_options = "HOLD_position") 
```

```{r}
data3 <- data2[data2$hipRL == "right",]
data3 <- data3[data2$treatment == "control",]
dt<-tidyr::pivot_longer(data3, c(9:10), names_to="Stat")
stats <- psych::describeBy(
  x = as.numeric(dt$value), group= dt$Stat,
  na.rm = TRUE, interp = TRUE, 
  quant= c(0.25,0.75), skew = TRUE, 
  digits = 2, mat = TRUE)
df4 <- data.frame(
  Treatment = c("Control", "Control"),
  Hip = c("Right", "Right"),
  Var = stats$group1, 
  `Sample Min` = stats$min, 
  `Sample Median` = stats$median, 
  `Sample Max` = stats$max, 
  SAM = stats$mean, 
  SAV = round((stats$sd)^2,3), 
  `Sample Skewness` = stats$skew)


knitr::kable(df, escape=F,
             caption = "Statistics by Variable for right hip",
             align = c('c', rep('r', 9))) %>%
kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
                          font_size = 9, latex_options = "HOLD_position") 
```


```{r}
dt<-tidyr::pivot_longer(data2[data2$treatment == "control",], c(9:10), names_to="Stat")
stats <- psych::describeBy(
  x = as.numeric(dt$value), group= dt$Stat,
  na.rm = TRUE, interp = TRUE, 
  quant= c(0.25,0.75), skew = TRUE, 
  digits = 2, mat = TRUE)
df8 <- data.frame(
  Treatment = c("Control", "Control"),
  Hip = c("Combined", "Combined"),
  Var = stats$group1, 
  `Sample Min` = stats$min, 
  `Sample Median` = stats$median, 
  `Sample Max` = stats$max, 
  SAM = stats$mean, 
  SAV = round((stats$sd)^2,3), 
  `Sample Skewness` = stats$skew)


# knitr::kable(df, escape=F,
#              caption = "Statistics by Variable for right hip",
#              align = c('c', rep('r', 9))) %>%
# kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
#                           font_size = 9, latex_options = "HOLD_position") 
```

```{r}

dt<-tidyr::pivot_longer(data2[data2$treatment == "treatment",], c(9:10), names_to="Stat")
stats <- psych::describeBy(
  x = as.numeric(dt$value), group= dt$Stat,
  na.rm = TRUE, interp = TRUE, 
  quant= c(0.25,0.75), skew = TRUE, 
  digits = 2, mat = TRUE)
df9 <- data.frame(
  Treatment = c("Treatment", "Treatment"),
  Hip = c("Combined", "Combined"),
  Var = stats$group1, 
  `Sample Min` = stats$min, 
  `Sample Median` = stats$median, 
  `Sample Max` = stats$max, 
  SAM = stats$mean, 
  SAV = round((stats$sd)^2,3), 
  `Sample Skewness` = stats$skew)


knitr::kable(df, escape=F,
             caption = "Statistics by Variable for right hip",
             align = c('c', rep('r', 9))) %>%
kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
                          font_size = 9, latex_options = "HOLD_position") 
```



```{r}
df <- rbind(df1, df2, df3, df8, df9, df4, df5, df6, df7)
knitr::kable(df, escape=F,
             caption = "Statistics by Variable",
             align = c('c', rep('r', 9))) %>%
kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
                          font_size = 9, latex_options = "HOLD_position") 
```
 Note: There is too much going on here, and it may be a good idea to either split these up or put them in a better order at some point


Below are some things that aren't necessarily necessary to show/present but just exploring how related flexion and rotation are related
```{r}
cor(data$flexionAfter, data$rotationAfter)
cor(data$flexionBefore, data$rotationBefore)
cor(data$flexionAfter-data$flexionBefore, data$rotationAfter-data$rotationBefore)
```
```{r}
plot(data$flexionAfter, data$rotationAfter)
plot(data$flexionBefore, data$rotationBefore)
plot(data$flexionAfter-data$flexionBefore, data$rotationAfter-data$rotationBefore)
```

##Added

calculate the difference on flexion and rotation before/after.



```{r}
data$flexiondiff=data$flexionAfter-data$flexionBefore
data$rotationdiff=data$rotationAfter-data$rotationBefore
plot(data$flexiondiff,type="l",col="red",xlab="observations including both hips", ylab="differences", main="before/after difference of flexion and rotation")
lines(data$rotationdiff,type="l",col="green")
abline(h=0)
legend("topleft",legend=c("flexion difference", "rotation difference"),
       col=c("red", "green"), lty=1:1)

# control group and different hips
# want to know if there is difference between right and left hips

par(mfrow=c(1,2))
plot(data[data$hipRL=="left"&data$treatment=='control',]$flexiondiff,type="o",xlab="12 Obs in Control", ylab="flexion difference",col="red",main="flexion difference",ylim=c(-15,55))
abline(h=0)
lines(data[data$hipRL=="right"&data$treatment=='control',]$flexiondiff,type="o",col="green")
legend("topleft",legend=c("left hip", "right hip"),
       col=c("red", "green"), lty=1:1)


plot(data[data$hipRL=="left"&data$treatment=='treatment',]$flexiondiff,type="o",xlab="27 Obs in Treatment", ylab="flexion difference",col="red",main="flexion difference",ylim=c(-15,55))
abline(h=0)
lines(data[data$hipRL=="right"&data$treatment=='treatment',]$flexiondiff,type="o",col="green")
legend("topleft",legend=c("left hip", "right hip"),
       col=c("red", "green"), lty=1:1)


par(mfrow=c(1,2))
plot(data[data$hipRL=="left"&data$treatment=='control',]$rotationdiff,type="o",xlab="12 Obs in Control", ylab="rotation difference",col="red",main="rotation difference",ylim=c(-10,22))
abline(h=0)
lines(data[data$hipRL=="right"&data$treatment=='control',]$rotationdiff,type="o",col="green")
legend("topright",legend=c("left hip", "right hip"),
       col=c("red", "green"), lty=1:1)


plot(data[data$hipRL=="left"&data$treatment=='treatment',]$rotationdiff,type="o",xlab="27 Obs in Treatment", ylab="rotation difference",col="red",main="rotation difference",ylim=c(-10,22))
abline(h=0)
lines(data[data$hipRL=="right"&data$treatment=='treatment',]$rotationdiff,type="o",col="green")
legend("bottomleft",legend=c("left hip", "right hip"),
       col=c("red", "green"), lty=1:1)


```
