---
title: "Sesame's Effect on Knowledge of Number and Letters"
author: "Jiahui Cheng, Leah Hunt, Ziling Zhang"
date: "September 21, 2020"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
---

```{r Front Matter, include=FALSE}

# clean up & set default chunk options
rm(list = ls())
knitr::opts_chunk$set(echo = FALSE)

# packages
library(tidyverse) 
library(ggplot2)
library(tidyr)
library(broom)
library(data.table)
library(caret)
sesame <- read.csv("~/BOAST_Summer_2020/Stat 470/sesame.csv",sep=",",header=T)
```


# Project Description

Sesame Street is an American educational children's television series that combines live action, sketch comedy, animation and puppetry. It is produced by Sesame Workshop (known as the Children's Television Workshop (CTW) ) . The program is known for its images communicated through the use of Jim Henson's Muppets, and includes short films, with humor and cultural references. 


The format of Sesame Street consists of a combination of commercial television production elements and 	techniques which have evolved to reflect the changes in American culture and the audience's viewing habits. The format of the show was designed to hold young children’s attention through action oriented, short duration presentations teaching specific preschool cognitive skills and some social skills. Each show was one hour and involved much repetition of concepts within and across shows. With the creation of Sesame Street, producers and writers of a children's television show used, for the first time, educational goals and a curriculum to shape its content. It was also the first time a show's educational effects were formally studied. 


In order to figure out whether to permit generalization to the populations of children of most interest to the producers, the Children's Television Workshop carried out an evaluation to the children who were sampled from five different sites in the US. Both before and after viewing the series the children were tested on a variety of cognitive variables (including knowledge on body parts, knowledge of letters, knowledge of forms, knowledge of numbers, knowledge of relations and knowledge of classifications skills) and collected the Sesame Street dataset.
  

As on the client's request, the report will focus on the effect of Sesame Street on improvement of test scores. While the study included six different score types, we have decided to focus on only two major categories that are commonly regarded metrics in intelligence: letters and numbers. This decision was made in light of the varying scales used for each test and limited knowledge on the general distribution of these scores and will be further justified within our report. The analysis will be done using a multilinear regression model fit individually for our two test types of interest with the response variable as the post test score and the predictors as the amount of sesame street watched as well as other variables that are expected to make a contribution to post test scores, such as the pre test score.



## Research Questions


**Question 1:**  Is Sesame Street effective in improving the knowledge of letters for the groups of interest?

**Question 2:**  Is Sesame Street effective in improving the knowledge of numbers for the groups of interest?


## Variables

We analyzed all of the variables collected in the study. Table 1 provides the name and a brief description of each variable. 

```{r fig.cap= "The table includes the no, name, description for each variable"}
# Variable table
variable<-data.frame(
  No = c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20),
  Variable=c("ID","SITE","SEX","AGE","VIEWCAT","SETTING","VIEWING","PREBODY",
             "PRELET","PREFORM","PRENUMB","PRERELAT","PRECLASF","POSTBODY",
             "POSTLET","POSTFORM","POSTNUMB","POSTREL","POSTCLAS","PEABODY"), 
  Description = 
    c("identification number",
      "Five different sampling sites coded as 1,2,3,4 or 5                          ",
      "Male=1, Female=2","In months",
      "1 for rarely watching the show to 4 if watching over 5 times a week",
      "        1 for home and coded as 2 for school",
      "1 for encouraged to view the show and 2 for not encouraged",
      "Pretest on knowledge about body parts",
      "       Pretest on knowledge about letters","Pretest on knowledge about forms",
      "       Pretest on knowledge about numbers",
      "Pretest on knowledge of relational terms",
      "       Pretest on knowledge of classification skills",
      "        Posttest knowledge on body parts",
      "      Posttest knowledge of letters",
      "Posttest knowledge of forms",
      "       Posttest knowledge of numbers","Posttest knowledge of relations",
      "       Posttest knowledge of classification skills","Mental age scores"))
knitr::kable(variable,
             caption = "Variable Attributes",
             align = c('c', rep('l', 9)),format="latex") %>%
kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
                          font_size = 9, latex_options = "HOLD_position")

```


# Exploratory Data Analysis (EDA)
```{r}
# Transform Data
sesame$sex <- sesame$sex - 1
sesameLongPre <- pivot_longer(sesame, 
                              cols = c("prebody", "prelet", "preform", 
                                       "prenumb", "prerelat", "preclasf"), 
                              names_to = "Pre", 
                              values_to = "ScorePre")
sesameLongPost <- pivot_longer(sesame, 
                               cols = c("postbody", "postlet",    "postform",  
                                        "postnumb",   "postrelat",  "postclasf"), 
                               names_to = "Post", values_to = "ScorePost")
sesameLong <- data.frame(Type = sesameLongPre$Pre, 
                         Post = sesameLongPost$Post, 
                         ScorePre = sesameLongPre$ScorePre, 
                         ScorePost = sesameLongPost$ScorePost, 
                         Category = as.factor(sesameLongPre$viewcat))
sesameLong$Type[sesameLong$Type=="prebody"] <- "Body Parts"
sesameLong$Type[sesameLong$Type=="prelet"] <- "Letters"
sesameLong$Type[sesameLong$Type=="preform"] <- "Forms"
sesameLong$Type[sesameLong$Type=="prenumb"] <- "Numbers"
sesameLong$Type[sesameLong$Type=="prerelat"] <- "Relationships"
sesameLong$Type[sesameLong$Type=="preclasf"] <- "Classification"
sesameTestScores <- data.frame(test = c(sesameLongPre$Pre, sesameLongPost$Post), 
                               score = c(sesameLongPre$ScorePre,
                                         sesameLongPost$ScorePost))
```

There were no clear errors or missing values in the dataset as presented. Table 2 shows the descriptive statistics for the relevant variables to our study with the index numbers corresponding to the positions from Table 1. Table 2 shows that the standard deviation values of postlet, prelet, postnumb and prenumb is much higher than other variables, which means that there are significant differences in the children’s knowledge of letters and numbers, both before and after viewing the Sesame Street television series. Thus, when combined with the commonality of these tests as a metric for knowledge, we justify placing our primary focus on score differences in letters and numbers as opposed to the other categories.

```{r fig.cap="The table displays descriptive statistics for variables, including minimum, maximum, median, mean value, standard deviation"}
# dt<-tidyr::pivot_longer(sesame, 1:23, names_to="Stat")
# stats <- psych::describeBy(
#   x = as.numeric(dt$value), group= dt$Stat,
#   na.rm = TRUE, interp = TRUE, 
#   quant= c(0.25,0.75), skew = TRUE, 
#   digits = 2, mat = TRUE)
# df <- data.frame(
#   Var = stats$group1, 
#   `Sample Min` = stats$min, 
#   `Sample Median` = stats$median, 
#   `Sample Max` = stats$max, 
#   SAM = stats$mean, 
#   SAV = round((stats$sd)^2,3), 
#   `Sample Skewness` = stats$skew)
# 
# knitr::kable(df, 
#              caption = "Statistics by Variable",
#              align = c('c', rep('r', 9)),format="latex") %>%
# kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),font_size = 9, latex_options = "HOLD_position")

dt<-tidyr::pivot_longer(sesame, 1:23, names_to="Stat")
stats <- psych::describeBy(
  x = as.numeric(dt$value), group= dt$Stat,
  na.rm = TRUE, interp = TRUE, 
  quant= c(0.25,0.75), skew = TRUE, 
  digits = 2, mat = TRUE)
df <- data.frame(
  Var = stats$group1, 
  `Sample Min` = stats$min, 
  `Sample Median` = stats$median, 
  `Sample Max` = stats$max, 
  SAM = stats$mean, 
  SAV = round((stats$sd)^2,3), 
  `Sample Skewness` = stats$skew)
df <- df[c(1,5:17, 20),]

knitr::kable(df, 
             caption = "Statistics by Variable",
             align = c('c', rep('r', 9))) %>%
kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
                          font_size = 9, latex_options = "HOLD_position")

```


Figure 1 shows the relationship between the pre- and post- test scores for each category at each viewing level. In the figure, we see that as viewing level increases, amount of improvement, shown on the graph as distance above the black line, appears to increase. We also see that the scoring patterns do vary somewhat between the different test types with scores like Classification and Relationships clustering around lower score numbers and scores like Letters and Numbers appearing as larger numbers, with often larger increases, and with much more variation, as we already saw. In spite of these wide variations, some of which is a result of the varying scoring scales of the different tests, it does appear that all of the variables at least loosely follow the trend of increasing improvement as the viewing category increases

```{r fig.height=3,fig.weight=4, fig.cap = "These charts plot pre- and post- test scores against one another by test type and viewing category. Points above the black line show improvement while points below the black line show a decrease in scores between the two tests."}
par(mfrow=c(2,2))
plot<-ggplot2::ggplot(aes(x = ScorePre, y=ScorePost, color= Type), data= sesameLong) +
        geom_point() + 
        geom_abline() +
        facet_wrap(facet = vars(Category)) +
        xlab("Prescores") + 
        ylab('Postscores') +
        ggtitle("Scoring Changes")+
        theme(axis.text = element_text(size=15),
              plot.title = element_text(size=15, face="bold",hjust=0.5),
              axis.title = element_text(size=15),
              panel.background = element_rect(fill = "white", color="black"),
              legend.text = element_text(size=14),
              legend.title = element_text(size=16)
        )

plot
```



```{r fig.height=3,fig.weight=10, fig.cap = "This figure shows the distributions of all of the test score types. Notice that they have different distributions, especially in letters and numbers where the posttest scores appear bimodal and the pretest scores appear skewed."}
plot<-ggplot2::ggplot(aes(x = score), data= sesameTestScores[sesameTestScores$test %in% c("prelet", "postlet", "prenumb", "postnumb"),]) +
        geom_histogram(bins=10) + 
        facet_wrap(facet = vars(test), ncol = 2) +
        xlab("Score") + 
        ylab('Number of children') +
        ggtitle("Distribution of Test Scores")+
        theme(axis.text = element_text(size=10),
              plot.title = element_text(size=10, face="bold",hjust=0.5),
              axis.title = element_text(size=10),
              panel.background = element_rect(fill = "white", color="black"),
              legend.text = element_text(size=14),
              legend.title = element_text(size=16)
        )

plot

```

When considering the analysis, we may question if variables, in particular gender and age, will affect our results. These variables are explored in Appendix 2 in Figure 5, which shows that while there are some variations in both variables between viewing categories, they are rather similar. 

Finally, take a look at the distributions of scores change before and after viewing the tv series. As is shown in Figure 2, the distributions of scores of letters and numbers before and after watching the show do appear to change. The post test scores do appear overall greater than the pretest scores, though in both cases the post test case takes on a mildly bimodal distribution relative to the positively skewed distribution of the pretest scores. A comparison of these distributions to those of the other test types from the study can be found in Figure 6 in Appendix 2. We see that the increasing nature is constant to these other distributions but the bimodal distribution of the post test scores are unique to numbers and letters. One possible theory for this difference is that of these skills, letters and numbers are least likely to be learned from sources beyond sesame street and thus the distributions change more drastically, though this theory is not adequately addressed by the data and thus is solely a possible explanation that we suggest.




# Statistical Analysis 

In this section, we use postlet and postnumb as response variable, respectively, and use other variables as explanatory variables. We will build two multiple linear regression models related to postlet and postnumb, separately. And do research on the relationships between postlet (postnumb)  with other varibales.

First, build a MLR model using postlet as response variable, prelet, viewcat, sex, age, the interaction term of sex and prelet, the interaction term of sex and viewcat , the interaction term of sex and age, the interaction term of age and prelet as explanatory variables. The summary of this MLR model is shown.

```{r fig.cap="Figure 4: The summary of this MLR model related to postlet"}
#mod <- lm(postlet ~ prelet + viewcat + sex + age + sex*prelet + sex*viewcat + sex*age + prelet*viewcat + age*viewcat + prelet*age + sex*prelet*viewcat + prelet*age*viewcat + sex*age*viewcat + prelet*age*sex, data = sesame)

mod <- lm(postlet ~ prelet + viewcat + sex + age + sex*prelet + sex*viewcat + sex*age, data = sesame)

# dataLet <- sesame[,c("prelet", "postlet", "sex", "peabody", "age", "viewcat")]
# dataLet$viewcat <- as.factor(dataLet$viewcat)
# dummies <- dummyVars(postlet~., data=dataLet)
# letters<-predict(dummies, newdata=dataLet)
# letters<-data.table(letters)
# letters$postlet<- dataLet$postlet
# letters <- letters[,-"viewcat.1"]
# mod <- lm(postlet~. + viewcat.2*age, data = letters)
out<-broom::tidy(mod)
knitr::kable(out)
```


The multiple linear regression model is:
\[
postlet = -11.69+ 0.88 prelet + 5.07 viewcat + 1.07 sex + 
\newline\]
\[
 + 0.033 prelet*sex -0.018 viewcat*sex – 0.0085 sex*age -0.0039  prelet*age
\]

From figure 3, it can be observed that the coefficient of viewcat is significant at significance level 0.001. Thus, we can conclude that the posttest on knowledge of children about letters is closely related to the viewing level. If a child often watched the show, then he or she can significantly get a increase in the knowledge about letters than those children who rarely watch the show.

Next, consider the postnumb as response variable. Build a MLR model using postnumb as response variable, prenumb, viewcat, sex, age, the interaction term of sex and prenumb, the interaction term of sex and viewcat, the interaction term of sex and age as explanatory variables. The summary of this MLR model is shown..

```{r fig.cap="Figure 6: The summary of this MLR model related to postnumb"}
#  mod2 <- lm(postlet ~ prelet + viewcat + sex + age + sex*prelet + 
# sex*viewcat + sex*age + sex*viewcat + age*prelet, data = sesame)

 mod2 <- lm(postnumb ~ prenumb + viewcat + sex + age + sex*prenumb + 
sex*viewcat + sex*age, data = sesame)

# datanumb <- sesame[,c("prenumb", "postnumb", "sex", "peabody", "age", "viewcat")]
# datanumb$viewcat <- as.factor(datanumb$viewcat)
# dummies <- dummyVars(postnumb~., data=datanumb)
# numbers<-predict(dummies, newdata=datanumb)
# numbers<-data.table(numbers)
# numbers$postnumb<- datanumb$postnumb
# numbers <- numbers[,-"viewcat.1"]
# numbers$sex <- numbers$sex - 1

#mod2 <- lm(postnumb~. + viewcat.2*sex + viewcat.3*sex, data = numbers)
out<-broom::tidy(mod2)
knitr::kable(out)
```

The multiple linear regression model is:

\[
postnumb = -5.05+ 0.35 prenumb + 3.28 viewcat + 2.90 sex 
  + 0.35 age 
  \newline\]
  \[+ 0.22 prenumb*sex -0.046viewcat*sex – 0.13sex*age
\]

From figure 3, it can be observed that the coefficients of prenumb, viewcat and the interaction trem of prenumb and sex are significant at significance level 0.05. Thus, we can conclude that the posttest on knowledge of children about numbers are closely related to the viewing level, the pretest on knowledge of children about numbers. If a child had a good knowledge before watching the show, and watched the show frequently, then he or she can significantly get an increase in the knowledge about numbers.




# Recommendations 

**Question 1:**  The first research question is about Sesame Street effective in improving the knowledge of letters for the groups of interest?

Our analysis has provided evidence that Sesame Street is effective in improving the knowledge of letters of children as improving the knowledge of letters is highly related to the viewing level. If a child watched Sesame Street more frequently, then he or she can get a higher improvement in the knowledge of letters

**Question 2:**  The second research question is about Sesame Street effective in improving the knowledge of numbers for the groups of interest?

The Sesame Street is effective in improving the knowledge of numbers of children. And improving the knowledge of numbers is related to the viewing level and the children’s knowledge of numbers before watching the show. If a child had a good knowledge of numbers and watched Sesame Street more frequently, then he or she can get a higher improvement in the knowledge of numbers.





# Resources 

The introduction about the Sesame Street show in the project description section is from Wikipedia: https://en.wikipedia.org/wiki/Sesame_Street


# Limitations

The analysis suggests that sesame street is effective on both number and letters abilities. However, the original experiment is done on many other aspects such as body parts. Thus, a further study on sesame's effect on other abilities is recommended.

Moreover, the analysis is built on some assumptions that the client might want to clarify. First, the selection of subjects are random, not out of interest, and that selections were done independently, for instance without taking groups of siblings. Selection bias might cause the study lack of generality. Second, the scoring rules of each category were not clearly described by the client. In particular, the analysis was unable to guage the significance of a particular score on any of the tests, which also prevented any form of normalizing or combining test scores. 

Overall, a suggestion to the client is to reevaluate the construction of the experiment and to make sure all assumptions above are made. We look forward to working with other potential effects of sesame street on children.


# Technical Appendix  

## Apendix 1: Assumption checks

Assumption checks for model1 with using postlet as response variable, prelet, viewcat, sex, age, the interaction term of sex and prelet, the interaction term of sex and viewcat , the interaction term of sex and age, the interaction term of age and prelet as explanatory variables.

we can observe from residuals vs fitted plot that the residuals are like white noise pattern. From normal Q-Q plot we can conclude that the residuals are normally distributed pattern. Thus, the multiple linear regression model of postlet does not violate assumptions for inference.

```{r fig.cap = "These plots show the assumptions tests for the letters model. We see that the normality assumption is fairly well met as the points on the normality plot are mostly within the error bands. The homoscedasticity assumption is fairly well met as the errors are mostly distributed randomly around 0. The final two plots show that there are no noticable patterns in the response or residuals by index, implying that the independence and linearity assumptions respectively are relatively well met."}
par(mfrow=c(2,2))
# mod <- lm(postlet ~ prelet + as.factor(viewcat) + sex + age + sex*prelet + sex*as.factor(viewcat) + sex*age + prelet*as.factor(viewcat) + age*as.factor(viewcat) + prelet*age + sex*prelet*as.factor(viewcat) + prelet*age*as.factor(viewcat) + sex*age*as.factor(viewcat) + prelet*age*sex, data = sesame)
a <- car::qqPlot(
  x = mod$residuals,
  distribution = "norm",
  envelope = 0.97,
  pch = 20
)
plot(mod, which = 1, pch = 19)
plot(mod$residuals)
plot(sesame$postlet)
```

Assumption checks for the second model that using postnumb as response variable, prenumb, viewcat, sex, age, the interaction term of sex and prenumb, the interaction term of sex and viewcat, the interaction term of sex and age as explanatory variables. 

We can observe from residuals vs fitted plot that the residuals are like white noise pattern. From normal Q-Q plot we can conclude that the residuals are normally distributed pattern. Thus, the multiple linear regression model of postnumb does not violate assumptions for inference

```{r fig.cap = "These plots show the assumptions tests for the letters model. We see that the normality plot has some mild devations from the error bands, but these deviations are mild enough that we will consider the normality assumption to be reasonably met. The homoscedasticity assumption is fairly well met as the errors are mostly distributed randomly around 0. The final two plots show that there are no noticable patterns in the response or residuals by index, implying that the independence and linearity assumptions respectively are relatively well met."}
par(mfrow=c(2,2))
# mod2 <- lm(postlet ~ prelet + viewcat + sex + age + sex*prelet + 
#              sex*viewcat + sex*age + sex*viewcat + age*prelet, data = sesame)
a <- car::qqPlot(
  x = mod$residuals,
  distribution = "norm",
  envelope = 0.97,
  pch = 20
)

plot(mod2, which = 1, pch = 19)
plot(mod2$residuals)
plot(sesame$postlet)
```


## Appendix 2: Additional Plots

```{r fig.height=3,fig.weight=4, fig.cap ="This plot shows the ages and genders of the participants in each viewing category. Notice that the age and gender distribution is relatively similar, though not exactly the same, for all four viewing categories."}
gender <- sesame$sex
gender[gender==0] <- "Male"
gender[gender==1] <- "Female"
gender <- as.factor(gender)
plot<-ggplot2::ggplot(aes(x = age, fill= gender), data= sesame) +
        geom_histogram(binwidth = 4) + 
        facet_wrap(facet = vars(viewcat)) +
        xlab("Age (months)") + 
        ylab('Number of children') +
        ggtitle("Distribution of Age and Gender")+
        theme(axis.text = element_text(size=15),
              plot.title = element_text(size=15, face="bold",hjust=0.5),
              axis.title = element_text(size=15),
              panel.background = element_rect(fill = "white", color="black"),
              legend.text = element_text(size=14),
              legend.title = element_text(size=16)
        )
plot

```

```{r fig.height=3,fig.weight=10, fig.cap = "This figure shows the distributions of all of the test score types. Notice that they have different distributions, especially in letters and numbers where the posttest scores appear bimodal and the pretest scores appear skewed."}
plot<-ggplot2::ggplot(aes(x = score), data= sesameTestScores) +
        geom_histogram(bins=10) + 
        facet_wrap(facet = vars(test), ncol = 6) +
        xlab("Score") + 
        ylab('Number of children') +
        ggtitle("Distribution of Test Scores")+
        theme(axis.text = element_text(size=10),
              plot.title = element_text(size=10, face="bold",hjust=0.5),
              axis.title = element_text(size=10),
              panel.background = element_rect(fill = "white", color="black"),
              legend.text = element_text(size=14),
              legend.title = element_text(size=16)
        )

plot

```


\newpage

### R Script
```{r codeAppendix, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}
```
